
\cleardoublepage

\addcontentsline{toc}{chapter}{Acknowledgements}

\begin{acknowledgements}

First and foremost, I would like to thank my supervisor, Professor Duncan Gillies, for his support and advice over the past three years. Without his help and patience during many difficult periods, this thesis would not have been possible. Duncan is not only an excellent mentor but also a researcher in its truest meaning. I feel privileged that I had the opportunity to work with and learn from him to be humane while striving for excellence. I would also like to thank Duncan for providing me the opportunity to help three MSc students in their dissertation as well as tutorial opportunities. 

I would like to take this opportunity to thank my thesis examiners, Professor Michael Stumpf and Professor David Gilbert, for agreeing to examine my work. Also, my second supervisor, Professor Daniel Rueckert, for examining my transfer report and his valuable advice.

Imperial College and Beit Trust funded my stay for three years in this expensive city (London) with \textit{Imperial College Deputy Rector's Scholarship} and \textit{Beit Fellowship for Scientific Research}. I am thankful from the bottom of my heart for providing me this opportunity. I would also like to express my gratitude to the Department of Computing for funding my travel costs both within the United Kingdom and internationally. Many thanks to the Zebra Housing Association, for providing me a place to stay at subsidised cost, without which I would certainly have missed the joy and privilege of living in central London so close to the university. 

I would like to thank all the denizens of Room-433 of Imperial College DOC (Huxley Building) for lifelong memories. This includes non-logicians Georgia, Alexei, Joao, Peter, Simon and logicians Uri, Mohammed, Jonathan, Mark, Nick and Clemens.
	
A huge thanks to all my friends, most importantly, Lennon and Alexei for their company, support and help during various ups and downs. I would also like to thank Orlando for many a stimulating discussion and help with filling gaps in my education. 

My thanks go to the BR forum members and multitudes of other websites that provided much needed distraction when I was not working. A big thank you, to the coffee shops in and around Imperial College for the caffeine, when I was going through the daunting task of writing up this thesis.

I am grateful to my parents for their unending love, support and encouragement. Without the values that they taught me, I would not be here. I thank all my family members for encouraging me, and bearing my mood changes all along.

Finally, and most importantly, my wife Saswati, for her endless love, support and encouragement throughout this journey. It would not have been possible without you.

\end{acknowledgements}
